In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-05 paper/main.tex
Updates from Overleaf

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged surface PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular enrichment and clearance. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

@@

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of molecular enrichment and clearance of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of model representation of the intrathecal tracer injection; 
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). To examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Methods). For the cardiac contribution, we compute
CSF pressure and flow fields at peak systolic blood inflow,
corresponding to a reduction in CSF space volume at a total rate of
$6\,$ml/s~\cite{baledent2014imaging, causemann2022human} in the SAS
and $0.31\,$ml/s across the lateral ventricle
surface\cite{vinje2019respiratory}. This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Human and rodent observations indicate that tracers concentrate in perivascular spaces surrounding the pial and subarachnoid vasculature \cite{zhang1990interrelationships,zhang1992directional, bedussi2017paravascular, mestre2018flow, bedussi2018paravascular, eide2024functional}. However, it remains unclear whether such enrichment patterns necessitate a structural barrier, such as a membrane with limited permeability, or if the patterns could result from enhanced flow or mixing in these areas alone. We therefore next investigate how the permeability of the interface between the PVS and the surrounding CSF affects tracer enrichment around the major arterial trunks (\Cref{fig:compartmentalization}A--B). To this end, we compare models with high and low  permeabilities, representing a highly permeable PVS-CSF interface (functional compartmentalization) and a less permeable interface (structural compartmentalization), respectively. For the low permeability model, we set the permeability to $3.8 \cdot 10^{-7}\,$m/s, consistent with previous estimates for the endfoot sheath surrounding penetrating arterioles \cite{koch2023estimates}, which we consider to be a lower bound for the surface PVS-CSF permeability. For the high permeability model, we increase this permeability by a factor of 100. We emphasize that, for both scenarios, we consider the high PVS flow regime as examined in the previous section.

@@

\subsection*{Enlarged PVSs in the SAS delay periarterial and intracranial molecular enrichment}

Idiopathic normal pressure hydrocephalus (iNPH), a dementia subtype,
is associated with enlarged PVS in the SAS and impaired periarterial
and parenchymal tracer transport~\cite{eide2024functional}. More
generally, enlarged PVSs are associated with a range of pathophysiogical
conditions and cognitive decline~\cite{bown2022physiology}. A key
question is thus whether enlarged PVSs alone leads to delayed
perivascular and intracranial tracer enrichment. To address this
question from a fluid and transport dynamics perspective, we ask our
in-silico model to predict the integrated effect of perivascular
dilation on PVS flow velocities and tracer enrichment patterns
(\Cref{fig:enlargement}A).

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement in the SAS reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS in the SAS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of molecular enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network, occurring with or
without pressure differences at the network ends. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

% Compartmentalization of the PVS
To what extent are surface PVSs in communication with the surrounding
SAS and to what extent are these structurally separated compartments?
Originally, Weller and coauthors~\cite{zhang1990interrelationships,
  zhang1992directional, weller2005microscopic} identified thin sheaths
of pial cells surrounding human surface arteries and penetrating
arterioles (but not veins or venules). On the other hand, Bakker and
coauthors report that the subarachnoid space, the cisterns, ventricles
and penetrating periarteriolar spaces form a continuous CSF
space~\cite{bedussi2017paravascular}. Thorne and colleagues study the
architectures of rat perivascular spaces in detail, emphasizing the
presence of openings (stomata) on the interface between the
vasculature and the CSF within the SAS~\cite{pizzo2018intrathecal,
  hannocks2018molecular}. Further, Mestre et
al~\cite{mestre2022periarteriolar} study the properties of pial
perivascular spaces in mice, and report of pial cells forming sheaths
for larger surface arteries and partially cover smaller surface
arteries, with higher coverage in ventral SAS regions. Our results
demonstrate that the PVSs may act as rapid transport pathways even
with only a partial barrier between the PVS and surrounding CSF, thus
contributing to quantifying the properties of the human PVS-CSF
interface~\cite{eide2024functional}. In this regard, we emphasize that
we report on the isolated effect of variations in permeability on
molecular transport, separated from any effects on fluid flow in the
PVS. Clearly, one may expect that the presence and/or properties of
such a barrier would affect also PVS flow. However, considering the
uncertainty associated with PVS flow characteristics in the human SAS
in general and the lack of computational models describing net PVS
flow in the presence of fluid exchange, this remains challenging to
quantify.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the
periarterial flow velocity is defined for each segment $\Lambda_i$ as
$\langle Q'_i \rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the
cross-section area of the PVS segment and $\langle Q_i' \rangle$ is
the estimated mean flow rate (see Supplementary information S1.4 for
more details). Two wall motion patterns are considered, yielding
$\hat{u}_a^c$ and $\hat{u}_a^v$ respectively: \emph{cardiac}
pulsations with $f = 1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$
(\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f
= 0.1$Hz, $\lambda = 0.02$m, and $\varepsilon = 10\%$
(\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for
the perivenous network. In the baseline model, we combine the
contributions from CSF production and cardiac peristalsis to set
$\hat{u}$ in~\eqref{eq:multi_transport} as $\hat{u}_a = \hat{u}_a^p +
\hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v = \hat{u}_v^p$ in
$\Lambda_v$. We remark that none of these models account for fluid
exchange across the lateral outer interface of the PVSs (over
$\Gamma_a, \Gamma_v$), while fluid entry and exit is allowed at the
end nodes of the PVS networks.

@@

    respectively, but with dilated PVS. }

>>>>

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of \mar{intracranial molecular enrichment and clearance}. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

@@

\subsection*{In-silico predictions of intracranial \mar{molecular enrichment and clearance} after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of \mar{molecular enrichment and clearance} of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of model representation of the intrathecal tracer injection; 
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). 
To examine the dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Supplementary information S1.3 for a complete description). 
To estimate the cardiac contribution, we compute the steady state
CSF pressure and flow fields at peak systolic blood inflow. Modelling the reduction 
in CSF space volume corresponding to peak systolic conditions, we assume a
total inflow rate of $6\,$ml/s~\cite{baledent2014imaging, causemann2022human} 
in the SAS across its outer surface and $0.31\,$ml/s across the lateral ventricle surface\cite{vinje2019respiratory}. This scenario sets up a pressure drop of 
$10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the spinal 
SAS and a maximum flow velocity of $19.8\,$cm/s (\Cref{fig:csf}F, J). 
Assuming a cardiac frequency of $1\,$Hz, we infer that this cardiac-induced 
pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Human and rodent observations indicate that tracers concentrate in perivascular spaces surrounding the pial and subarachnoid vasculature \cite{zhang1990interrelationships,zhang1992directional, bedussi2017paravascular, mestre2018flow, bedussi2018paravascular, eide2024functional}. However, it remains unclear whether such enrichment patterns necessitate a structural barrier, such as a membrane with limited permeability, or if the patterns could result from enhanced flow or mixing in these areas alone. We therefore next investigate how the permeability of the interface between the PVS and the surrounding CSF affects tracer enrichment around the major arterial trunks (\Cref{fig:compartmentalization}A--B). To this end, we compare models with high and low  permeabilities, representing a highly permeable PVS-CSF interface (functional compartmentalization) and a less permeable interface (structural compartmentalization), respectively. For the low permeability model, we set the permeability to $3.8 \cdot 10^{-7}\,$m/s, consistent with previous estimates for the endfoot sheath surrounding penetrating arterioles \cite{koch2023estimates}, which we consider to be a lower bound for the surface PVS-CSF permeability. For the high permeability model, we increase this permeability by a factor of 100. For both scenarios, we consider the high PVS flow regime as examined in the previous section.

@@

\subsection*{Enlarged PVSs delay periarterial and intracranial molecular enrichment}

Enlarged PVSs are associated with impaired periarterial tracer
transport and more generally with cognitive decline as well as
certain subtypes of
dementia~\cite{bown2022physiology,eide2024functional}. A key
question is thus whether enlarged PVSs alone leads to delayed perivascular
and intracranial tracer enrichment. To address this question from a
fluid and transport dynamics perspective, we ask our in-silico model
to predict the integrated effect of perivascular dilation on PVS flow
velocities and tracer enrichment patterns (\Cref{fig:enlargement}A).

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of \mar{molecular
enrichment and clearance} in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs. A key observation is that the balance between CSF production and
intracranial pulsatility significantly shapes the large-scale features
of molecular enrichment, with e.g.~ventricular tracer reflux after
intrathecal injection in the absence of CSF production, and preference
towards anterior brain regions with reduced pulsatility. Indeed, the
simulated distribution patterns cover a substantial range of the
individual variations observed clinically \cite{ringstad2018brain}. In
terms of perivascular pathways, we find that even moderate CSF flow on
the order of 0.1--1.0 \textmu m/s in the surface PVSs results in
earlier tracer enrichment around major cerebral arteries. This effect
is more pronounced -- with up to a three-hour difference in time of
arrival -- with mean PVS flow rates of approximately 10 \textmu m/s. Our
models of CSF flow in perivascular networks, based on first principles
and asymptotic analysis, predict net flow of such magnitudes and both
antegrade and retrograde flow within the perivascular network,
occurring with or without pressure differences at the network
ends. The PVS may thus function as highways facilitating rapid
transport, even in the absence of structural barriers, though a sharp
concentration difference between the PVS and surrounding tissues is
unlikely without such barriers. Finally, dilated PVS will cause a
substantial reduction in cardiac- and vasomotion-driven flow
velocities, which in turn impedes perivascular transport.

@@

% Compartmentalization of the PVS
To what extent are surface PVSs in communication with the surrounding
SAS and to what extent are these structurally separated compartments?
Originally, Weller and coauthors~\cite{zhang1990interrelationships,
  zhang1992directional, weller2005microscopic} identified thin sheaths
of pial cells surrounding human surface arteries and penetrating
arterioles (but not veins or venules). On the other hand, Bakker and
coauthors report that the subarachnoid space, the cisterns, ventricles
and penetrating periarteriolar spaces form a continuous CSF
space~\cite{bedussi2017paravascular}. Thorne and colleagues study the
architectures of rat perivascular spaces in detail, emphasizing the
presence of openings (stomata) on the interface between the
vasculature and the CSF within the SAS~\cite{pizzo2018intrathecal,
  hannocks2018molecular}. Further, Mestre et
al~\cite{mestre2022periarteriolar} study the properties of pial
perivascular spaces in mice, and report of pial cells forming sheaths
for larger surface arteries and partially cover smaller surface
arteries, with higher coverage in ventral SAS regions. Our results
demonstrate that the PVSs may act as rapid transport pathways even
with only a partial barrier between the PVS and surrounding CSF, thus
contributing to quantifying the properties of the human PVS-CSF
interface~\cite{eide2024functional}.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see Supplementary information S1.4 for more details). Two
wall motion patterns are considered, yielding $\hat{u}_a^c$ and
$\hat{u}_a^v$ respectively: \emph{cardiac} pulsations with $f =
1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$ (\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ (\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for the perivenous network. In the baseline model, we combine the contributions from CSF production and cardiac peristalsis to set $\hat{u}$ in~\eqref{eq:multi_transport} as
$\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v
= \hat{u}_v^p$ in $\Lambda_v$.

@@

    respectively, but with dilated PVS.}
